Gene expression profiling of reproductive meristem types in early rice inflorescences by laser microdissection

### 2016-02-03 ms.Rmd
Comments Martin Vero Israr

Thomas W.R. Harrop^§,1^, Israr Ud Din^¶,^[^1], Veronica Gregis^¶^, Michela Osnato^¶^, Stefan Jouannic^§^, Hélène Adam^§,2^ and Martin Kater^¶,^[^2]

@@

| ^§^ Institut de Recherche pour le Développement, UMR DIADE, 911 Avenue Agropolis, 34394 Montpellier, France.
| ^¶^ Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.

@@

Several previous transcriptomic studies have measured gene expression in whole rice inflorescences at later stages of development than those described here [*e.g.* @Wang:2010im; @Sato:2011eh; @Sharma:2012kq; @Khanday:2013bz; @Jiang:2014im]. Another study used microarray experiments at earlier stages to highlight the importance of transcription factors in panicle development [@Furutani:2006hx]. LMD has also been used for precise control of developmental stage during the collection of whole panicle sections for microarrays [@Kobayashi:2012fi]. This identified three MADS genes, which are members of the *SQUA*-like (*FUL*-like) clade, that are co-expressed with *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), and a quadruple knockdown of these four genes resulted in defects in inflorescence development [@Kobayashi:2012fi]. All four genes were detected in all stages in the RNAseq dataset (Data S1), confirming their expression in reproductive meristems. In contrast to previous transcriptomic studies, only the meristematic regions of young panicles were collected by LMD for the RNAseq dataset presented in this article, to restrict the measurement of gene expression to those tissues.

@@

The molecular mechanisms controlling the meristematic activities that guide panicle development in rice remain largely uncharacterized. The dataset presented here addresses this by describing both the transcriptome profiles associated with specific meristem identities and the changes in gene expression that occur during the early stages of panicle development. A marked expression switch was evident between apical and axillary meristems, followed by gradual changes during transitions between PBM and SM. This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which undergo transition to determinate SMs. The latter process may be comparable to the gradual meristem maturation observed during tomato inflorescence development [@Park:2012dg], rather than to the continual production of determinate floral meristems characteristic of the Arabidopsis inflorescence meristem. Promising uncharacterized genes were identified, notable examples including transcription factor genes with expression patterns similar to known regulators of panicle architecture and putative hormone-related such as the auxin-responsive genes. Functional studies of these genes will reveal the mechanics of the regulatory networks that establish gene expression patterns in meristems and their control by hormone signalling. Genes with expression specific to certain meristem types were also identified, which may prove valuable as marker genes, for example in developmental studies of mutants affected in determination of reproductive meristem identity. Combined with current technology for the creation of mutants, this dataset promises to accelerate research on the molecular mechanisms controlling panicle development, a vital component of efforts to achieve the sustainable increase in grain yield required to address population growth.

>>>>

Thomas W.R. Harrop^§,1^, Israr Ud Din^¶,†,^[^1], Veronica Gregis^¶^, Michela Osnato^¶,‡^, Stefan Jouannic^§^, Hélène Adam^§,2^ and Martin Kater^¶,†,^[^2]

@@

| ^§^ Institut de Recherche pour le Développement, UMR DIADE, 911 Avenue Agropolis, 34394 Montpellier, France.
| ^¶^ Dipartimento di Bioscienze, Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.
| ^†^ Dipartimento di Scienze Farmacologiche e Biomolecolari, Università degli Studi di Milano, Via Celoria 26, 20133 Milano, Italy.
| ^‡^ Present address: Centre for Research in Agricultural Genomics, Campus UAB, 08193 Cerdanyola, Spain.

@@

Several previous transcriptomic studies have measured gene expression in whole rice inflorescences at later stages of development than those described here [*e.g.* @Wang:2010im; @Sato:2011eh; @Sharma:2012kq; @Khanday:2013bz; @Jiang:2014im]. Another study used microarray experiments at earlier stages to highlight the importance of transcription factors in panicle development [@Furutani:2006hx]. LMD has also been used for precise control of developmental stage during the collection of whole panicle sections for microarrays [@Kobayashi:2012fi]. This identified three MADS genes, which are members of the *SQUA*-like (*FUL*-like) clade, that are co-expressed with *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), and a quadruple knockdown of these four genes resulted in defects in inflorescence development [@Kobayashi:2012fi]. All four genes were detected in all stages in the RNAseq dataset presented in this article (Data S1), confirming their expression in reproductive meristems. In contrast to previous transcriptomic datasets, only the meristematic regions of young panicles were collected by LMD for the RNAseq analysis described here, to restrict the measurement of gene expression to those tissues.

@@

The molecular mechanisms controlling the meristematic activities that guide panicle development in rice remain largely uncharacterized. The dataset presented here addresses this by describing both the transcriptome profiles associated with specific meristem identities and the changes in gene expression that occur during the early stages of panicle development. A marked expression switch was evident between apical and axillary meristems, followed by gradual changes during transitions between PBM and SM. This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which undergo transition to determinate SMs. The latter process may be comparable to the gradual meristem maturation observed during tomato inflorescence development [@Park:2012dg], rather than to the continual production of determinate floral meristems characteristic of the Arabidopsis inflorescence meristem. Promising uncharacterized genes were identified, notable examples including transcription factor genes with expression patterns similar to known regulators of panicle architecture, and putative hormone-related genes such as the auxin-responsive genes. Functional studies of these genes will reveal the mechanics of the regulatory networks that establish gene expression patterns in meristems and their control by hormone signalling. Genes with expression specific to certain meristem types were also identified, which may prove valuable as marker genes, for example in developmental studies of mutants affected in determination of reproductive meristem identity. Combined with current technology for the creation of mutants such as CRISPR/Cas9 genome editing, this dataset promises to accelerate research on the molecular mechanisms controlling panicle development, which is a vital for efforts to achieve the sustainable increase in grain yield required to address population growth.
